### 2021-06-07 article/article.tex
Version 0.3.4, cos_exp_m1 implementation

\section{Implementation and architecture}
	\subsection{Mathematical methods}
		\subsubsection{Quantum mechanics background}
			In general, \texttt{spinsim} solves the Schroedinger equation,

			\begin{align}
				\frac{\mathrm{d}}{\mathrm{d}t}\psi(t) &= -i2\pi H(t)\psi(t),\label{eq:schroedinger}
			\end{align}

			where \(i^2 = -1\), the quantum state \(\psi(t) \in \mathbb{C}^N\) a time dependent, \(N\) dimensional, unit complex vector, and the Hamiltonian \(H(t) \in \mathbb{C}^{N \times N}\) is a time dependent \(N \times N\) complex Hermitian matrix. Here \(N\) is the number spin levels in the quantum system, so spin half is the \(N = 2\) case, and spin one refers to the \(N = 3\) case. Furthermore, we have chosen for the Hamiltonian \(H(t)\) to have the physical dimension of frequency, rather than the more typical dimension of energy. To elaborate, we represent energies \(E\) of the system as their equivalent frequencies \(f\) from the Planck relation

@@

			where \(\hbar\) is (reduced) Planck's constant (or equivalently, setting 
			\(\hbar = 1\)). In addition, rather than being represented by standard coordinates in \(\mathbb{C}^{N \times N}\), in \texttt{spinsim} the Hamiltonian \(H(t)\) is instead represented with respect to a choice of basis for the corresponding Lie Algebra, \(\mathfrak{su}(N)\). For example, when set to spin half mode, the \texttt{spinsim} package solves the time dependent Schroedinger equation of the form

			\begin{align}
				\frac{\mathrm{d}}{\mathrm{d}t}\psi(t) = -i 2\pi (f_x(t) J_x + f_y(t) J_y + f_z(t) J_z) \psi(t),
			\end{align}

@@

			\begin{align}
				J_x &= \frac12\begin{pmatrix}
					0 & 1 \\
					1 & 0
				\end{pmatrix},
				&J_y &= \frac12\begin{pmatrix}
					0 & -i \\
					i &  0
				\end{pmatrix},
				&\textrm{and }J_z &= \frac12\begin{pmatrix}
					1 &  0 \\
					0 & -1
				\end{pmatrix}.\label{eq:spin_half_operators}
			\end{align}

			The fields \(f\) that represents the time dependent Hamiltonian \(H(t)\), are the collection of energy functions \(f_x(t), f_y(t), f_z(t)\), with time \(t\) and \(f\) with physical dimension of frequency that control the dynamics of the system. The user must define a method that returns a sample of these field functions when a sampling time is input. In physical terms, these functions could represent the \(x,y,z\) components of a magnetic field applied to a magnetically spin one BEC or NVC.

@@

			The matrices \(J_x, J_y, J_z\) are regular spin operators, and \(Q\) is a quadrupole operator. Note that \(Q\) is proportional to \(Q_{zz}\) as defined by Hamley et al \cite{hamley_spin-nematic_2012}, and \(Q_0\) as defined by Di et al \cite{di_dipolequadrupole_2010}.

			Frequently when one calculates the times series of the quantum state \(\psi(t)\), they are also interested in its expected spin projection \(\left\langle J\right\rangle\)(t). It is given by the vector

			\begin{align}
				\left\langle J\right\rangle(t) &= \begin{pmatrix}
					\psi(t)^\dagger J_x \psi(t)\\
					\psi(t)^\dagger J_y \psi(t)\\
					\psi(t)^\dagger J_z \psi(t)
				\end{pmatrix},
			\end{align}

@@

		\subsubsection{Unitary time evolution and the Magnus expansion}
			Given that \(\psi(t)\) is a unit vector, it is possible to write \(\psi(t)\) in terms of a unitary transformation \(U(t, t_0)\) of the state \(\psi(t_0)\), for any time \(t_0\). In other words, for any \(t_0,t \in \mathbb{R}\), there is a unitary transformation \(U(t, t_0)\) such that

@@

		\subsubsection{Exponentiator}
			From Equation \eqref{eq:cf4_implementation}, evaluating matrix exponentials is a core part of the integration algorithm. However, the \texttt{spinsim} package uses the user provided field functions \(f(t)\) as a vector representation of the Hamiltonian of the spin system being simulated, rather than generic matrix elements in an array. This allows for implementation of matrix exponentiation that takes advantage of this representation. For all implementations of exponentiation, the \texttt{spinsim} computes the matrix exponential

@@

		\subsubsection{Descretisation and parallelisation}
			Consider quantising time with
			\begin{align}
				t_k &= t_0 + \mathrm{D}t\cdot k,
			\end{align}

@@

		\subsubsection{Rotating frame}
			If the rotating frame option is selected, the \(U_k\) are first calculated within a rotating frame of reference as \(U^r_k\), which, in some situations, reduces the size of the field functions used in the calculation, increasing accuracy. The rotation speed of the rotating frame is calculated locally for each parallel time step \(U_k\), and only for rotations around the \(z\) axis. The rotating from field functions \(f^r_x, f^r_y, f^r_z,\) and \(f^r_q\) are related to the field function from the user input via

@@

	\subsection{Software architecture}
		\subsubsection{Integrator architecture}
			The integrator in the \texttt{spinsim} package calls a \texttt{numba.cuda.jit()}ed kernel to be run on a \emph{Cuda} capable \emph{Nvidia} GPU in parallel, with a different thread being allocated to each of the \(U_k\). This returns when each of the \(U_k\) have been evaluated.

@@

		\subsubsection{Compilation of integrator}
			The \texttt{spinsim} integrator is constructed and compiled just in time, using \texttt{numba.cuda.jit()}. The particular device functions used are not predetermined, but are instead chosen based on user input to decide a closure. This structure has multiple advantages. First, the field function \(f\) is provided by the user as a plain python function (that must be \texttt{numba.cuda.jit()} compatible). This allows users to define \(f\) in a way that compiles and executes fast, does not put many restrictions on the form of the function, and returns the accurate results of analytic functions (compared to the errors seen in interpolation). Compiling the simulator also allows the user to set meta parameters, and choose the features they want to use, in a way that does not require experience with the \texttt{numba.cuda} library. This was especially useful for running benchmarks comparing old integration methods to the new ones, like CF4. The default settings should be optimal for most users, although tuning the values of \emph{Cuda} meta parameters \texttt{max\_registers} and \texttt{threads\_per\_block} could improve performance for GPUs with a differing number of registers and \emph{Cuda} cores to the mobile \emph{GeForce GTX 1070} mainly used in testing here. Finally, just in time compilation also allows the user to select a target device other than \emph{Cuda} for compilation, so the simulator can run, using the same algorithm, on a multicore CPU in parallel instead of a GPU, if the user so chooses.

@@

\section{Quality control}
	\subsection{Benchmarks}
		\subsubsection{Evaluation of accuracy}
		All accuracy benchmarks were run in \emph{Cuda} mode, on the desktop computer with the \emph{Ryzen 7 5800X} and \emph{GeForce RTX 3080}, which from the tests in Figure \ref{fig:benchmark_device_aggregate} are the fastest CPU and GPU from the devices tested. Note that these benchmarks do not take into account the amount of time required to JIT compile simulation code for the first time (order of seconds), as we want to look at this in the limit of running many simulations while sweeping through a parameter that differs in each one.

@@

		\subsubsection{Comparison to alternatives}
			We ran the same error (using Equation \eqref{eq:error}) and execution time benchmarks on some alternative packages to compare \texttt{spinsim}'s performance to theirs. The packages compared were:
			\begin{itemize}
				\item \texttt{spinsim} running on the \emph{Cuda} device, using the CF4 integrator and the rotating frame mode, which were the highest performing options when comparing \texttt{spinsim} integration methods.
				\item \texttt{NDSolve} from the \emph{Mathematica} \cite{wolfram_research_inc_mathematica_2020} software. This was chosen as it is popular with our lab group for simulating magnetometry experiments.
				\item \texttt{scipy.integrate.ivp\_solve()} from the \emph{python} library \texttt{SciPy} \cite{virtanen_scipy_2020}. This was chosen as a generic solver from within the python ecosystem.
				\item We had also planned to benchmark against \texttt{qutip.sesolve()}, a solver in the popular quantum mechanics \emph{python} library, \texttt{QuTip} \cite{johansson_qutip_2013}. However, due to a known bug with the library’s dependencies, this was not installable on Windows 10, the operating system being used for testing, and so benchmarks for it could not be run.
			\end{itemize}

@@

		\subsubsection{Parallelisation performance}
			Once the algorithm behind \texttt{spinsim} was developed, we wanted to check its execution speed while running on various devices. Part of this test was also to quantify the speed increase of parallelisation by comparing execution speeds on highly parallel devices (being GPUs), compared highly procedural devices (being CPUs). Speed benchmarks were performed using \texttt{sense.sim.benchmark}, by comparing evaluation speed of typical spin one sensing experiments on different devices. This is shown in Figure \ref{fig:benchmark_device_aggregate}. The integration code was compiled by \texttt{numba} for multicore CPUs, CPUs running single threaded, and \emph{Nvidia Cuda} compatible GPUs, and run on different models of each of them. These test devices are given in Table \ref{tab:devices}.

>>>>

\section{Implementation and architecture}
	% \subsection{Mathematical methods}
	\subsection{Quantum mechanics background}
		\texttt{spinsim} solves the time-dependent Schr\"{o}dinger equation

		\begin{align}
			i\hbar\frac{\mathrm{d}\psi(t)}{\mathrm{d}t} &= H(t)\psi(t),\label{eq:schroedinger}
		\end{align}

		where the quantum state \(\psi(t) \in \mathbb{C}^N\) is assumed normalised, and the Hamiltonian \(H(t) \in \mathbb{C}^{N \times N}\) is Hermitian. Here \(N\) is the number of levels in the quantum system. Often we are considering systems with spin \(S\). Here spin-half is the \(N = 2\) case, and spin-one is \(N = 3\). We set \(\hbar = 1\), so that the Hamiltonian has units

		% Furthermore, we have chosen for the Hamiltonian \(H(t)\) to have the physical dimension of frequency, rather than the more typical dimension of energy. To elaborate, we represent energies \(E\) of the system as their equivalent frequencies \(f\) from the Planck relation

@@

		% where \(\hbar\) is (reduced) Planck's constant (or equivalently, setting 
		% \(\hbar = 1\)). In addition, rather than being represented by standard coordinates in \(\mathbb{C}^{N \times N}\), in \texttt{spinsim} the Hamiltonian \(H(t)\) is instead represented with respect to a choice of basis for the corresponding Lie Algebra, \(\mathfrak{su}(N)\). For example, when set to spin half mode, the \texttt{spinsim} package solves the time dependent Schroedinger equation of the form
		Rather than ... parametrizing the problem ... we consider time varying real coefficients in a linear combination of fixed operators,

		\begin{align}
			H(t) &= \sum_{j = 1}^{N^2 - 1} a_j(t) A_j.
		\end{align}

		It is well known that ... spin in magnetic field has ... . We exclude the identity ... energy zero point.

		In the spin one case there is no standard basis ... . Choices include Gell Mann, quadrupole basis, ... . In general ... Lie algebra ... mathfrak ... admits a basis ... .
		% \begin{align}
		% 	\frac{\mathrm{d}}{\mathrm{d}t}\psi(t) = -i 2\pi (f_x(t) J_x + f_y(t) J_y + f_z(t) J_z) \psi(t),
		% \end{align}

@@

		DONT WRITE OUT, sigma
		\begin{align}
			J_x &= \frac12\begin{pmatrix}
				0 & 1 \\
				1 & 0
			\end{pmatrix},
			&J_y &= \frac12\begin{pmatrix}
				0 & -i \\
				i &  0
			\end{pmatrix},
			&\textrm{and }J_z &= \frac12\begin{pmatrix}
				1 &  0 \\
				0 & -1
			\end{pmatrix}.\label{eq:spin_half_operators}
		\end{align}

		Call omega, no implementation detail
		% The fields \(f\) that represents the time dependent Hamiltonian \(H(t)\), are the collection of energy functions \(f_x(t), f_y(t), f_z(t)\), with time \(t\) and \(f\) with physical dimension of frequency that control the dynamics of the system. The user must define a method that returns a sample of these field functions when a sampling time is input. In physical terms, these functions could represent the \(x,y,z\) components of a magnetic field applied to a magnetically spin one BEC or NVC.

@@

		The matrices \(J_x, J_y, J_z\) are regular spin operators, and \(Q\) is a quadrupole operator. Note that \(Q\) is proportional to \(Q_{zz}\) as defined by Hamley et al \cite{hamley_spin-nematic_2012}, and \(Q_0\) as defined by Di et al \cite{di_dipolequadrupole_2010}.
		INLINE Q DIAG. WHY WE'VE PICKED THESE FOUR. For ground state spin one atoms in a magnetic field, ... Breit Rabi ... . COULD BE READILY EXTENDED TO INCLUDE FURTHER ELEMENTS OR TO HIGHER SPINS.

		% Too far out of scope
		% Alternative: integrate Bloch equations 3 coupled real equations
		% fast and parallel function from wavefunctions
		Frequently when one calculates the times series of the quantum state \(\psi(t)\), they are also interested in its expected spin projection \(\left\langle J\right\rangle\)(t). It is given by the vector

		% Inline
		\begin{align}
			\left\langle J\right\rangle(t) &= \begin{pmatrix}
				\psi(t)^\dagger J_x \psi(t)\\
				\psi(t)^\dagger J_y \psi(t)\\
				\psi(t)^\dagger J_z \psi(t)
			\end{pmatrix},
		\end{align}

@@

	\subsection{Unitary time evolution and the Magnus expansion}
		Given that \(\psi(t)\) is a unit vector, it is possible to write \(\psi(t)\) in terms of a unitary transformation \(U(t, t_0)\) of the state \(\psi(t_0)\), for any time \(t_0\). In other words, for any \(t_0,t \in \mathbb{R}\), there is a unitary transformation \(U(t, t_0)\) such that

@@

	\subsection{Exponentiator}
		From Equation \eqref{eq:cf4_implementation}, evaluating matrix exponentials is a core part of the integration algorithm. However, the \texttt{spinsim} package uses the user provided field functions \(f(t)\) as a vector representation of the Hamiltonian of the spin system being simulated, rather than generic matrix elements in an array. This allows for implementation of matrix exponentiation that takes advantage of this representation. For all implementations of exponentiation, the \texttt{spinsim} computes the matrix exponential

@@

	\subsection{Discretisation and parallelisation}
		Consider quantising time with
		\begin{align}
			t_k &= t_0 + \mathrm{D}t\cdot k,
		\end{align}

@@

	\subsection{Rotating frame}
		If the rotating frame option is selected, the \(U_k\) are first calculated within a rotating frame of reference as \(U^r_k\), which, in some situations, reduces the size of the field functions used in the calculation, increasing accuracy. The rotation speed of the rotating frame is calculated locally for each parallel time step \(U_k\), and only for rotations around the \(z\) axis. The rotating from field functions \(f^r_x, f^r_y, f^r_z,\) and \(f^r_q\) are related to the field function from the user input via

@@

% \subsection{Software architecture}
	\subsection{Integrator architecture}
		The integrator in the \texttt{spinsim} package calls a \texttt{numba.cuda.jit()}ed kernel to be run on a \emph{Cuda} capable \emph{Nvidia} GPU in parallel, with a different thread being allocated to each of the \(U_k\). This returns when each of the \(U_k\) have been evaluated.

@@

	\subsection{Compilation of integrator}
		The \texttt{spinsim} integrator is constructed and compiled just in time, using \texttt{numba.cuda.jit()}. The particular device functions used are not predetermined, but are instead chosen based on user input to decide a closure. This structure has multiple advantages. First, the field function \(f\) is provided by the user as a plain python function (that must be \texttt{numba.cuda.jit()} compatible). This allows users to define \(f\) in a way that compiles and executes fast, does not put many restrictions on the form of the function, and returns the accurate results of analytic functions (compared to the errors seen in interpolation). Compiling the simulator also allows the user to set meta parameters, and choose the features they want to use, in a way that does not require experience with the \texttt{numba.cuda} library. This was especially useful for running benchmarks comparing old integration methods to the new ones, like CF4. The default settings should be optimal for most users, although tuning the values of \emph{Cuda} meta parameters \texttt{max\_registers} and \texttt{threads\_per\_block} could improve performance for GPUs with a differing number of registers and \emph{Cuda} cores to the mobile \emph{GeForce GTX 1070} mainly used in testing here. Finally, just in time compilation also allows the user to select a target device other than \emph{Cuda} for compilation, so the simulator can run, using the same algorithm, on a multicore CPU in parallel instead of a GPU, if the user so chooses.

@@

\section{Quality control}
	% \subsection{Benchmarks}
	\subsection{Evaluation of accuracy}
	All accuracy benchmarks were run in \emph{Cuda} mode, on the desktop computer with the \emph{Ryzen 7 5800X} and \emph{GeForce RTX 3080}, which from the tests in Figure \ref{fig:benchmark_device_aggregate} are the fastest CPU and GPU from the devices tested. Note that these benchmarks do not take into account the amount of time required to JIT compile simulation code for the first time (order of seconds), as we want to look at this in the limit of running many simulations while sweeping through a parameter that differs in each one.

@@

	\subsection{Comparison to alternatives}
		We ran the same error (using Equation \eqref{eq:error}) and execution time benchmarks on some alternative packages to compare \texttt{spinsim}'s performance to theirs. The packages compared were:
		\begin{itemize}
			\item \texttt{spinsim} running on the \emph{Cuda} device, using the CF4 integrator and the rotating frame mode, which were the highest performing options when comparing \texttt{spinsim} integration methods.
			\item \texttt{NDSolve} from the \emph{Mathematica} \cite{wolfram_research_inc_mathematica_2020} software. This was chosen as it is popular with our lab group for simulating magnetometry experiments.
			\item \texttt{scipy.integrate.ivp\_solve()} from the \emph{python} library \texttt{SciPy} \cite{virtanen_scipy_2020}. This was chosen as a generic solver from within the python ecosystem.
			\item We had also planned to benchmark against \texttt{qutip.sesolve()}, a solver in the popular quantum mechanics \emph{python} library, \texttt{QuTip} \cite{johansson_qutip_2013}. However, due to a known bug with the library’s dependencies, this was not installable on Windows 10, the operating system being used for testing, and so benchmarks for it could not be run.
		\end{itemize}

@@

	\subsection{Parallelisation performance}
		Once the algorithm behind \texttt{spinsim} was developed, we wanted to check its execution speed while running on various devices. Part of this test was also to quantify the speed increase of parallelisation by comparing execution speeds on highly parallel devices (being GPUs), compared highly procedural devices (being CPUs). Speed benchmarks were performed using \texttt{sense.sim.benchmark}, by comparing evaluation speed of typical spin one sensing experiments on different devices. This is shown in Figure \ref{fig:benchmark_device_aggregate}. The integration code was compiled by \texttt{numba} for multicore CPUs, CPUs running single threaded, and \emph{Nvidia Cuda} compatible GPUs, and run on different models of each of them. These test devices are given in Table \ref{tab:devices}.
